---
title: State of the Ocean report
date: "`r Sys.Date()`"
author: Pieter Provoost
output: (function(...) {
  rmdformats::robobook(toc_depth = 4, pandoc_args = c("+RTS", "-K2000m", "-RTS"), ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

## Dependencies

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(sf)
library(robis)
library(rnaturalearth)
library(stringr)
library(cartomisc)
library(arrow)
library(nngeo)
library(RColorBrewer)

theme_simple <- theme(
  axis.title.x = element_blank(),
  panel.grid.major.y = element_line(color = "#cccccc"),
  panel.background = element_blank(),
  panel.grid.major.x = element_blank(),
  panel.grid.minor.x = element_blank(),
  axis.line.x = element_line(color = "#000000"),
  legend.position = "bottom",
  legend.title = element_blank(),
  legend.key = element_blank()
)
```

## Background spatial data

```{r message=FALSE, warning=FALSE}
countries <- ne_countries(returnclass = "sf")
land <- landr::get_land_polygons(simplified = TRUE) %>%
  st_simplify(dTolerance = 30000) %>%
  filter(!st_is_empty(geometry))
```

## Oceans and seas

```{r message=FALSE, warning=FALSE}
if (!file.exists("data/goas.shp")) {
  download.file("http://geo.vliz.be/geoserver/MarineRegions/ows?service=WFS&version=1.0.0&request=GetFeature&typeName=MarineRegions:goas&maxFeatures=50&outputFormat=SHAPE-ZIP", "data/goas.zip")
  unzip("data/goas.zip", exdir = "data")
}

goas <- st_read("data/goas.shp", options = "ENCODING=UTF-8", quiet = TRUE)

sf_use_s2(FALSE)

goas_simplified <- goas %>%
  st_simplify(dTolerance = 0.1) %>%
  nngeo::st_remove_holes() %>%
  select(name) %>%
  st_set_crs(NA) %>%
  st_segmentize(1) %>%
  st_set_crs(4326)
```

## Type localities dataset

```{r message=FALSE, warning=FALSE}
localities <- occurrence(datasetid = "b74b429a-4052-4f5b-bff3-fe0b5a2e8669") %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

breaks <- c(1800, 1900, 1950, 2000, 2100)
labels <- head(paste0(breaks, " - ", breaks[2:length(breaks)]), -1)
labels <- str_replace(labels, "2100", "")
labels <- str_replace(labels, "1800", "")

localities <- localities %>%
  mutate(period = cut(date_year, breaks = breaks, labels = labels))

ggplot() +
  geom_sf(data = land, fill = NA, color = "#000000", size = 0.1) +
  geom_sf(data = localities %>% filter(!is.na(period)), size = 0.3, color = "#000000") +
  coord_sf(crs = "ESRI:54030") +
  theme_void() +
  facet_wrap(~period, ncol = 2)

ggsave("output/type_localities_map.png", height = 6, width = 10, dpi = 400, bg = "white")
```

Version with Oceans and Seas polygons:

```{r message=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = goas_simplified, aes(fill = name), size = 0) +
  geom_sf(data = land, fill = "#ffffff", color = "#000000", size = 0.1) +
  geom_sf(data = localities %>% filter(!is.na(period)), size = 0.3, color = "#000000") +
  scale_fill_brewer(palette = "Paired") +
  coord_sf(crs = "ESRI:54030") +
  theme_void() +
  theme(legend.position = "none") +
  facet_wrap(~period, ncol = 2)

ggsave("output/type_localities_map_oceans.png", height = 6, width = 10, dpi = 400, bg = "white")
```

Join type localities with Oceans and Seas data shapefile:

```{r message=FALSE, warning=FALSE}
localities$goas <- goas_simplified$name[st_nearest_feature(localities, goas_simplified, check_crs = TRUE)]

ggplot() +
  geom_sf(data = land, fill = NA, color = "#000000", size = 0.1) +
  geom_sf(data = localities %>% filter(!is.na(period)), aes(color = goas), size = 0.3) +
  coord_sf(crs = "ESRI:54030") +
  theme_void() +
  scale_color_brewer(palette = "Paired") +
  theme(legend.position = "none") +
  facet_wrap(~period, ncol = 2)

ggsave("output/type_localities_map_color_nolegend.png", height = 6, width = 10, dpi = 400, bg = "white")

ggplot() +
  geom_sf(data = land, fill = NA, color = "#000000", size = 0.1) +
  geom_sf(data = localities %>% filter(!is.na(period)), aes(color = goas), size = 0.3) +
  coord_sf(crs = "ESRI:54030") +
  theme_void() +
  scale_color_brewer(palette = "Paired") +
  guides(color = guide_legend(title = "Oceans and Seas")) +
  facet_wrap(~period, ncol = 2)

ggsave("output/type_localities_map_color.png", height = 4, width = 9, dpi = 400, bg = "white")
```

Frequency table:

```{r message=FALSE, warning=FALSE}
localities_cleaned <- localities %>%
  as.data.frame() %>%
  filter(!is.na(period))

stats <- round(table(localities_cleaned$goas, localities_cleaned$period) / nrow(localities_cleaned) * 100, digits = 2)
stats %>% knitr::kable()
```

```{r message=FALSE, warning=FALSE}
stats <- localities_cleaned %>%
  group_by(period, goas) %>%
  summarize(n = n())

ggplot(stats) +
  geom_bar(aes(x = period, y = n, fill = goas), stat = "identity") +
  scale_fill_brewer(palette = "Paired") +
  guides(fill = guide_legend(title = "Oceans and Seas"))

ggsave("output/type_localities_barplot.png", height = 4, width = 7, dpi = 300, bg = "white", scale = 1.5)

ggplot(stats) +
  geom_bar(aes(x = period, y = n, fill = goas), stat = "identity", position = "dodge") +
  scale_fill_brewer(palette = "Paired") +
  guides(fill = guide_legend(title = "Oceans and Seas"))

ggsave("output/type_localities_barplot_dodge.png", height = 4, width = 7, dpi = 300, bg = "white", scale = 1.5)
```

## Occurrence statistics

Load all OBIS occurrences from S3:

```{r message=FALSE, warning=FALSE}
space <- S3FileSystem$create(
  anonymous = TRUE,
  scheme = "https",
  endpoint_override = "ams3.digitaloceanspaces.com"
)

occ <- open_dataset(space$path("obis-datasets/exports/obis_20220208.parquet")) %>%
  select(date_year, date_mid, species, decimalLongitude, decimalLatitude) %>%
  as_tibble()
```

Join with Oceans and Seas:

```{r message=FALSE, warning=FALSE}
if (!file.exists("data/occ_joined.dat")) {
  occ_coords <- occ %>%
    distinct(decimalLongitude, decimalLatitude) %>%
    st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, remove = FALSE)
  occ_joined <- st_join(occ_coords, goas_simplified, join = st_within)
  remove(occ_coords)
  save(occ_joined, file = "data/occ_joined.dat")
} else {
  load("data/occ_joined.dat")
}

occ <- occ %>%
  left_join(occ_joined %>% as.data.frame() %>% select(decimalLongitude, decimalLatitude, name), by = c("decimalLongitude", "decimalLatitude"))
remove(occ_joined)
```

Verify the results by sampling some points:

```{r message=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = land, fill = NA, color = "#000000", size = 0.1) +
  geom_sf(data = occ %>% sample_n(10000) %>% st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326), aes(color = name), size = 0.3) +
  coord_sf(crs = "ESRI:54030") +
  theme_void() +
  scale_color_brewer(palette = "Paired") +
  guides(color = guide_legend(title = "Oceans and Seas"))
```

Statistics:

```{r message=FALSE, warning=FALSE}
stats <- occ %>%
  group_by(name) %>%
  summarize(records = n(), species = length(unique(species)))

stats
```

Statistics over time:

```{r message=FALSE, warning=FALSE}
stats <- occ %>%
  filter(date_year >= 1900 & !is.na(name)) %>%
  group_by(name, date_year) %>%
  summarize(records = n(), species = length(unique(species))) %>%
  ungroup() %>%
  tidyr::complete(name, date_year)

ggplot(stats %>% filter(!is.na(name))) +
  geom_smooth(aes(x = date_year, y = records, color = name), fill = NA, method = "gam") +
  geom_point(aes(x = date_year, y = records, color = name)) +
  annotate(
    geom = "text", x = min(stats$date_year), y = max(stats$records, na.rm = TRUE), 
    label = "(a)", hjust = 0, vjust = 0, size = 7
  ) +
  scale_color_brewer(palette = "Paired") +
  guides(color = guide_legend(title = "Oceans and Seas")) +
  scale_y_continuous(trans = "log10", name = "Number of records") +
  scale_x_continuous(breaks = scales::breaks_pretty(5)) +
  theme_simple

ggsave("output/records_time.png", height = 5, width = 7, dpi = 300, bg = "white", scale = 1.5)

ggplot(stats %>% filter(!is.na(name))) +
  geom_bar(aes(x = date_year, y = records, fill = name), stat = "identity", size = 0, width = 1) +
  scale_color_brewer(palette = "Paired") +
  scale_y_continuous(trans = "log10") +
  scale_x_continuous(breaks = scales::breaks_pretty(5)) +
  facet_wrap(~name) +
  theme_simple

ggplot(stats %>% filter(!is.na(name))) +
  geom_smooth(aes(x = date_year, y = species, color = name), fill = NA, method = "gam") +
  geom_point(aes(x = date_year, y = species, color = name)) +
  annotate(
    geom = "text", x = min(stats$date_year), y = max(stats$species, na.rm = TRUE), 
    label = "(b)", hjust = 0, vjust = 0, size = 7
  ) +
  scale_color_brewer(palette = "Paired") +
  scale_x_continuous(breaks = scales::breaks_pretty(5)) +
  scale_y_continuous(name = "Number of species") +
  theme_simple

ggsave("output/species_time.png", height = 5, width = 7, dpi = 300, bg = "white", scale = 1.5)

ggplot(stats %>% filter(!is.na(name))) +
  geom_smooth(aes(x = date_year, y = species, color = name), fill = NA, method = "gam") +
  geom_point(aes(x = date_year, y = species, color = name)) +
  scale_color_brewer(palette = "Paired") +
  scale_y_continuous(trans = "log10") +
  theme_simple

ggsave("output/species_time_log.png", height = 5, width = 7, dpi = 300, bg = "white", scale = 1.5)

pal <- rev(brewer.pal(11, "RdBu"))

ggplot(stats %>% filter(!is.na(name))) +
  geom_tile(aes(x = date_year, y = 1, fill = species)) +
  scale_fill_gradientn(colors = pal, na.value = pal[1]) +
  facet_wrap(~name, ncol = 3) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.y = element_blank(), axis.line.y = element_blank(), axis.title = element_blank())

ggsave("output/species_stripes.png", height = 4, width = 7, dpi = 300, bg = "white", scale = 1.5)

ggplot(stats %>% filter(!is.na(name))) +
  geom_tile(aes(x = date_year, y = 1, fill = species)) +
  scale_fill_gradientn(colors = pal, na.value = pal[1], trans = "log10") +
  facet_wrap(~name, ncol = 3) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.text.y = element_blank(), axis.line.y = element_blank(), axis.title = element_blank())

ggsave("output/species_stripes_log.png", height = 4, width = 7, dpi = 300, bg = "white", scale = 1.5)
```

Combined OBIS records trend:

```{r message=FALSE, warning=FALSE}
stats <- occ %>%
  filter(date_year >= 1900 & !is.na(name)) %>%
  group_by(date_year) %>%
  summarize(records = n(), species = length(unique(species))) %>%
  ungroup()

ggplot(stats) +
  geom_bar(aes(x = date_year, y = records), stat = "identity", width = 1, fill = "#DE8F6E") +
  scale_y_continuous() +
  scale_x_continuous(breaks = scales::breaks_pretty(5)) +
  theme_simple
```

Cumulative species:

```{r message=FALSE, warning=FALSE}
stats_cumulative <- occ %>%
  filter(!is.na(name) & !is.na(species) & !is.na(date_year)) %>%
  group_by(name, species) %>%
  summarize(date_year = min(date_year)) %>%
  group_by(name, date_year) %>%
  summarize(species = n()) %>%
  arrange(name, date_year) %>%
  mutate(cumulative_species = cumsum(species))

ggplot(stats_cumulative) +
  geom_line(aes(x = date_year, y = cumulative_species, color = name), size = 1) +
  scale_color_brewer(palette = "Paired") +
  guides(color = guide_legend(title = "Oceans and Seas")) +
  xlim(c(1900, 2025)) +
  theme_simple

ggsave("output/species_cumulative.png", height = 4, width = 7, dpi = 300, bg = "white", scale = 1.5)
```

## Correcting for sampling effort

```{r message=FALSE, warning=FALSE}
create_grid <- function() {
  b <- st_as_sfc(st_bbox(c(xmin = -180, xmax = 180, ymin = -90, ymax = 90))) %>%
    st_set_crs(4326) %>%
    st_transform(3410)
  h <- st_make_grid(b, cellsize = 100000, square = FALSE) %>%
    st_intersection(b) %>%
    st_as_sf() %>%
    mutate(cell = row_number())
  i <- as.numeric(st_area(h)) > max(as.numeric(st_area(h))) * 0.99
  h[i,] %>% st_transform(4326)
}

if (!file.exists("data/occ_joined_hex.dat")) {
  grid <- create_grid()
  occ_joined_hex <- occ %>%
    distinct(decimalLongitude, decimalLatitude) %>%
    st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, remove = FALSE) %>%
    st_join(grid, join = st_within)
  save(occ_joined_hex, file = "data/occ_joined_hex.dat")
} else {
  load("data/occ_joined_hex.dat")
}

occ <- occ %>%
  left_join(occ_joined_hex %>% as.data.frame() %>% select(decimalLongitude, decimalLatitude, cell), by = c("decimalLongitude", "decimalLatitude")) %>%
  mutate(date_month = lubridate::month(as.POSIXct(as.numeric(date_mid), origin = "1970-01-01")))

remove(occ_joined_hex)
```

```{r message=FALSE, warning=FALSE}
sampling_effort <- occ %>%
  filter(!is.na(name) & !is.na(cell) & date_year >= 1900) %>%
  group_by(date_year, name, cell, date_month) %>%
  summarize() %>%
  group_by(date_year, name) %>%
  summarize(effort = n())

ggplot(sampling_effort) +
  geom_smooth(aes(x = date_year, y = effort, color = name), fill = NA, method = "gam") +
  geom_point(aes(x = date_year, y = effort, color = name)) +
  scale_color_brewer(palette = "Paired") +
  guides(color = guide_legend(title = "Oceans and Seas")) +
  scale_y_continuous(trans = "log10") +
  scale_x_continuous(breaks = scales::breaks_pretty(5)) +
  theme_simple

stats_cumulative <- stats_cumulative %>%
  left_join(sampling_effort, by = c("date_year", "name")) %>%
  filter(date_year >= 1900)

ggplot(stats_cumulative) +
  geom_smooth(aes(x = date_year, y = species / effort, color = name), fill = NA, method = "gam") +
  geom_point(aes(x = date_year, y = species / effort, color = name)) +
  annotate(
    geom = "text", x = min(stats_cumulative$date_year), y = max(stats_cumulative$species / stats_cumulative$effort, na.rm = TRUE), 
    label = "(c)", hjust = 0, vjust = 0, size = 7
  ) +
  scale_color_brewer(palette = "Paired") +
  guides(color = guide_legend(title = "Oceans and Seas")) +
  scale_y_continuous(trans = "log10", name = "Species added, corrected for sampling effort") +
  scale_x_continuous(breaks = scales::breaks_pretty(5)) +
  theme_simple

ggsave("output/species_new_corrected_log.png", height = 4, width = 7, dpi = 300, bg = "white", scale = 1.5)
```
